Implementing and Evaluating Space Efficient Algorithms for Detecting Large Neighbourhoods in Graph Streams

### 2020-05-13 paper/src/paper.tex
Submitted

\usepackage[ruled, vlined, linesnumbered]{algorithm2e}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{changepage}
\usepackage{enumerate}
\usepackage{enumitem}
\usepackage{fancyhdr} % Header
\usepackage{float} % [H] for figure
\usepackage{graphicx}
\usepackage{subfig}
\usepackage{tikz}

@@

@@

\renewcommand\thetable{\arabic{table}} % figure numbering strategy
\renewcommand{\thefigure}{\arabic{chapter}.\arabic{section}.\arabic{figure}}

\pagenumbering{roman}

@@

% Title
\title{Implementing and Evaluating Space Efficient Algorithms for Detecting Large Neighbourhoods in Graph Streams}
\author{Dom Hutchinson}
\date{\today}
\maketitle
\newpage

\chapter*{Declaration}

@@

\par The \texttt{Neighbourhood Detection Problem} is a problem in graph theory which tasks one with finding a vertex in the graph which has a certain number of neighbours and to then return a subset of these neighbours. This is a trivial problem to solve in theory, but in practice is hard to solve within a time frame which makes the implementation practical. A major reason for this is that many approaches to solving the problem require a lot of space and so quickly overflow a computer's RAM, slowing the execution significantly down. This common space inefficiency is becoming increasingly problematic since the move towards Big Data.

\par In this paper I discuss and implement the two algorithms presented in \cite{orig} which offer space efficient solutions for \texttt{neighbourhood detection} for two types of graph stream. During the implementation I discuss and evaluate several methods for improving performance of these algorithms, focusing on space-efficiency in order to maximise the potential size of graphs the implementations are effective on.

\par The final implementation of the proposed algorithm for insertion-only streams proves very promising with it able to find large neighbourhoods in a couple of minutes for a graph with 30 million edges. The algorithm for insertion-deletion streams proves much more difficult to evaluate due to restrictions incurred from the implementation of $L_0$ samplers.

@@

@@

%TODO why big data is important
% The commercial movement towards data science and the desire to find relationships has lead to big data....
\par The movement towards Big Data has lead to ever larger data sets which traditional offline algorithms cannot work with efficiently due to their high memory-requirements. As the size of data-sets is increasing at a much greater rate than standard RAM capacities means that virtual memory has to be used for these algorithms, significantly increasing their run-times.
\par The \textit{Data Stream Model} was introduced as a method for coping with these large data sets. In the \textit{Data Stream Model} instructions are received sequentially, as a `stream', and generally random access is not allowed. \textit{Graph Streams} are a common implementation of this model, where instructions describe updates to the edge set of a graph and are typically ordered as a time-series. This allows for a dynamic representation of a graph, which typical \texttt{adjacency list} implementations do not.
%TODO citation=https://en.wikipedia.org/wiki/Streaming_algorithm#Data_stream_model
\par Streaming algorithms are a class of algorithms, first formalised in 1996 \cite{TODO}, to work within the \textit{Data Stream Model}. Streaming algorithms are typically designed to work in $polylog(n)$, where $n$ is the number of vertices in the case of graph streams. Streaming algorithms are designed to work in as few passes of the stream as possible, preferable only a single-pass is used as this means the stream can easily be extended without the algorithm having to start over. Due to the limited space streaming algorithms are allocated, they are often produce an approximation of a solution in order to improve space efficiency. When implementing a streaming algorithm the trade-off between accuracy and space-usage needs to be considered.

\par Graphs are used to represent relationships between objects and are popular in data science. A graph stream is a sequence of instructions which describe how to construct a graph. The sequential nature of a graph stream allows it to represent the evolution of a graph over time, which adjacency matrices cannot. This makes graph streams ideal for representing changing networks such as a social or computer network where objects can connect and then disconnect.

\par A basic problem in graph theory is \texttt{High Degree Detection} where you want to find the vertices with the greatest degree. This problem can be trivially and efficiently solved by counting the number of edges incident to each vertex and then returning any vertices of sufficient degree. Solving \texttt{High Degree Detection} allows for the identification of the most influential vertices in a graph, however you cannot make any inferences about the nature of this influence. In a social network it is possible for someone to have lots of followers (\ie be of high degree), but for all these followers to be bot accounts meaning the account actually has zero influence.

@@

\par\texttt{Neighbourhood Detection} is the natural set up from \texttt{High Degree Detection}, it tasks you with finding vertices of high degree \underline{and} a subset of their neighbours. Solving \texttt{Neighbourhood Detection} allows for inferences to be made about the influence a vertex has. Extending the social network example, \texttt{Neighbourhood Detection} allows for analysis of the followers of a popular account which can be used for targeted marketing campaigns.

@@

\par In this paper I shall discuss two streaming algorithms which solving \texttt{Neighbourhood Detection} in $polylog(n)$ space. This is achieved by using a series of samplers in order to reduce the set of edges which need to be analysed. This means that there is a possibility of failure and this possibility increases as less edges are sampled, but there is a trade-off here with space efficiency.
%TODO move this sentence & rewrite
Using graph streams is good here as it allows for more space efficient algorithms but due to time-ordered nature of most graph streams their is an added difficulty that instructions do not arrive in the most optimal order for solving \texttt{Neighbourhood Detection}.

@@

\section{Objectives}
%TODO expand
The objectives of this project are to:
\begin{itemize}
  \item Understand the problem posed in \cite{orig} and its context in practical situtations.
	\item Implement and understand the algorithms proposed in \cite{orig}.
  \item Implement na\"ive algorithms and test their performance against the proposed.
  \item Show that na\"ive algorithms have a limit to the size of graph they can work on and that it is much lower than for the proposed algorithms.
  \item Investigate the practicalities of the samplers used in these algorithms.
  \item Discuss and test alterations to the logic and implementation of the proposed algorithms, and the subroutines they rely upon, in order to improve the performance of the proposed algorithm.
  \item Discuss the trade-offs in practicalities these alterations have, namely between speed and space usage.
	\item Use large graphs from real-world scenarios to evaluate the practicalities of these implementations on real-world data sets.
  \item Discuss how graphs from different real-world sources (\ie not social networks) might perform on these implementation and how the implementations could be changed to accomodate for them.
\end{itemize}

@@

Since the formalisation of the data streaming model in 1996 much research has been done into solving problems using this model. Here I give an overview of previous works in this field.\newline

@@

\par\noindent\textbf{Minimum Distance between two nodes}
\par A $\alpha$-\textit{spanner} subgraph $H$ of a graph $G=(V,E)$ is a subgraph where $\forall\ u,v\in V$ $d_G(u,v)\leq d_H(u,v)\leq\alpha\cdot d_G(u,v)$ where $d_G(\cdot,\cdot)$ is the length of the shortest path between the given nodes in $G$. Spanners are a space-efficient structure for evaluating the length of shortest path between any two nodes in a graph. A $(2t-1)$-spanner requires at most $\mathcal{O}(n^{1+\frac1t})$ edges need to be stored \cite{edgesOfSpanner}.
\par Baswana \cite{spanner} presents an single-pass streaming algorithm for unweighted insertion-only graph streams which runs in $\mathcal{O}(m)$ time. This is an amortised run-time of $\mathcal{O}(1)$ per edge, a natural limit for performance.\newline

\par\noindent\textbf{Random walks}
\par A random walk from $v_0\in V$ in a graph is any random sequence of vertices $v_0,v_1,v_2,\dots$ where $v_i$ is randomly selected from the neighbourhood of $v_{i-1}$. The distribution of $v_t$ can be assessed.
\par Das Sarma et al. \cite{RandomWalks} present an algorithm for approximating the distribution of $v_t$ in the context of estmating the \textit{PageRank} of a vertex. They present an algorithm which can perform $\frac{n}{l}$ indendepent random walks of length $l$ in $o(n)$ space and $o(l)$ passes. They use this algorithm to sample random walks of different lengths in order to estimate the distribution of $v_t$ and thus estimate \textit{PageRank}.\newline

@@

\section{Structure}
Having provide some background to the problem, the structure of this project is as follows. In \texttt{Chapter 1} I discuss and define general ideas, technologies and techniques which are used throughout this project. In \texttt{Chapter 2} I discuss, implement and evaluate the algorithm proposed in \cite{orig} for insertion-only graph streams. In \texttt{Chapter 3} I discuss, implement and evaluate the algorithm proposed in \cite{orig} for insertion-deletion graph streams. And, in \texttt{Chapter 4} review what was achieved in this project and provide some ideas for possible future work to build upon those achievements.

@@

\par A graph $G=(V,E)$ is said to be a \textit{Bipartite Graph} if its vertex set $V$ can be partitioned into two disjoint subsets $A,B$ and the every edges connects a vertex in $A$ to a vertex in $B$. Formally, $\exists\ A,B\subseteq V$ st $A\cup B=V,\ A\cap B=\emptyset$ and $\forall\ (a,b)\in E$ we have that $a\in A,b\in B$. \textit{Bipartite Graphs} are denoted as $G=(A\cup B,E)$.

@@

\par The \textit{Neighbourhood} of a vertex $v$ is the set of vertices which share an edge with, $N_v=\{u:(u,v)\in E\}$. The \textit{Degree} of a vertex $v$ is the size of its neighbourhood, $\delta_v=|N_v|$. A vertex $v$ is \textit{$s$-sparse} if its degree is less than, or equal to, $s$.
\par The \textit{Edge Vector} of a vertex $v$ of length $|V|$ where each index gives the weight of the edge from $v$ to the vertex associated with that index. Typically a weight of 0 is stored if no edge exists between $v$ and a specific vertex. The graphs used in this project are unweighted so the values in the edge vector are booleans describing whether, or not, an edge exists. These edge vectors can be stored as bitstrings.
\par \textit{Graph Streams} are an unbounded sequence of instructions $\{e_1,e_2,\dots\}$ which describe how to construct the graph. These instructions $e_i$ describe modifications to the edges of a graph. The vertex set of the described graph is inferred from the instructions. In this project we are only concerned with two types of graph stream. %TODO something about the unbounded part??
\begin{itemize}
	\item \textit{Insertion-Only Streams} where each instruction inserts a new edge to the graph. It is assumed that no edges are inserted multiple times. For the unweighted, undirected graphs used in this project insertion-only streams are a list of the edges in the graph and can be read in any order. Here the instructions have the form $e_i=(u,v)$ where $u,v\in V$ are the endpoints of the edge being inserted.
	\item \textit{Insertion-Deletion Streams} where each instruction either adds a new edge to the graph or removes an existing edge from the graph. It is assumed that no edges are inserted multiple times, nor is an edge removed if it is not currently in the graph. For the unweighted, undirected graphs used in this project instructions of insertion-deletion streams take the form $e_i=(\Delta,u,v)$ where $u,v\in V$ are the endpoints of the edge and $\Delta$ is a boolean defining whether $e_i$ is an insertion or deletion instruction. The order of instructions is clearly important for insertion-deletion streams.
\end{itemize}
Other versions of graph streams exist which allow for changes to the weights of edges and directed edges, but these are out of scope for this project.

@@

\par An \textit{$\alpha$-Approximation Algorithm} is an algorithm which returns a result within an $\alpha$ factor to the optimal solution to the problem. Algorithms which solve the \texttt{Neighbourhood Detection Problem} are $\frac1c$-approximation algorithms since they only return $d/c$ of the neighbours of a vertex which we know to have at least $d$ neighbours.

@@

\par A family of hash functions $\mathcal{H}=\{h:X\to Y\}$ is described as begin \textit{$k$-Wise Independent} if for any $k$ distinct elements of $X$ $\{v_1,\dots,v_k\}$ and any, not necessarily distinct, elements of $Y$ $\{a_1,\dots,a_k\}$ if $h$ is chosen uniformly at random from $\mathcal{H}$ then $$\mathbb{P}(h(v_1)=a_1\wedge\dots\wedge h(v_k)=a_k)=\frac1{|Y|^2}$$
\section{Technologies}
%\begin{itemize}
%	\item C++
%	\begin{itemize}
%		\item Low level
%		\item Memory manipulation
%		\item Faster
%		\item Perhaps something about why not java or python etc.
%	\end{itemize}
%	\item Standard library online
%\end{itemize}
For the implementations in this project I chose to use C++, specifically \texttt{C++11}. I initially considered using either python or C++, as they both work in the object-orientated paradigm and I am proficient in both. I chose C++ as it allows for more control with memory management and typically has faster run-times. For this project the memory management is more important as I am looking to evaluate the space-efficiency of algorithms. The faster run-times is important when evaluating the practicalities of the algorithms.
\par I limited myself to using the \textit{C++ Standard Library} \texttt{std} as the underlying implementations of this library are well document. This was particularly important for the abstract data types I used as I could check what the underlying implementations and adjust my evaluation accordingly. When implementing randomness I used the \texttt{<random>} module of \texttt{std} and seeded the generators with the current time so that each run would have a different generator.

@@

During evaluation three performance metrics were measured: success-rate, time-taken, and, space-used. These are all interconnected and come with their own trade-offs. Typically, the changes to space-used and time-taken when different strategies are implemented occur in the same direction, but at different rates. During most tests the success-rate was near perfect, but the other metrics were significantly higher as it took longer for a solution to be found.

\par In real world scenarios we are most interested analysing the neighbourhoods of the most influential vertices. For this reason $d$ is set to the maximum degree in the graph during all tests. Further, the greater the proportion of a neighbourhood returned the better the inferences made are so I focussed testing on low values of $c$, namely $c\in[1,20]$.

\par For this project I wanted some graphs generated from the real-world. I found a collection of large graph streams, from different applications, in the \texttt{Stanford Network Analysis Project} (\texttt{SNAP}) datasets collection \cite{SNAP}. I chose to use the graphs generated from social networks as I found that application most motivating, but this choice is essentially arbitrary. The graph streams from \cite{SNAP} were all insertion-only so I wrote a utility program which created insertion-deletion streams from them by using a Bernoulli random variable to decide whether to delete an edge immediately after inserting it. \textbf{Table 1} provides details about the graphs used during evaluation.

@@

@@

\par The reservoir $R$ has an invariant that at any point in time it contains a uniform sample of the vertices whose degrees are known to be greater than $d_1$. This invariant is proved as \textbf{Lemma 2.1}, first I shall describe how the reservoir is maintained by controlling how vertices are inserted. The first time the degree counter for a vertex $v$ surpasses $d_1$, $v$ is considered for insertion into the \textit{reservoir}. There are two cases
\begin{itemize}
	\item If the reservoir is \underline{not} full (\ie $|R|<s$) then $v$ is inserted into the reservoir.
	\item Otherwise, a Bernoulli random variable with probability $p=\frac{x}s$ is used, where $x$ is the number of vertices known to have degree at least $d_1$. If this random variable succeeds then pick, uniformly at random, a vertex $u$ currently in the reservoir and replace it with $v$. All the edges in $E$ which are incident to $u$, and no other vertex in the reservoir, are removed from $E$.
\end{itemize}

Whenever an edge $(u,v)$ is encountered and $u$ is in the reservoir and the number of edges stored in $E$ incident to $u$ is less than $d_2$ then $(u,v)$ is inserted into $(u,v)$. %TODO rewrite this
\begin{adjustwidth}{.3cm}{}\fbox{\parbox{\textwidth}{

@@

Hence by mathematical induction, $R$ contains a uniform sample of the vertices of sufficient degree at all points in time.$\hfill\square$
}}\end{adjustwidth}\vspace{.3cm}
%TODO adding termination condition? Maybe add during implementation

@@

\texttt{Neighbourhood Detection} can be solved using \texttt{degree-based reservoir sampling} by setting $d_2=\frac{d}c$. The question remains as to what to set $d_1$ and $s$ to in order to achieve a high probability of success. \textbf{Algorithm 2} is the space-efficient algorithm proposed in \cite{orig} for solving \texttt{Neighbourhood-Detection} for insertion-only streams.

\begin{algorithm}
	\caption{One-pass $c$-Approximation Insertion-Only Streaming Algorithm for $\mathtt{Neighbourhood\ Detection}$}
	\SetKwInOut{Require}{require}
	\Require{Space $s$, degree bound $d$.}
	$s\leftarrow\lceil\log(n)\cdot n^{\frac1c}\rceil$\\
	\For{$i\in[0,c-1]$\text{\textbf{ in parallel}}} {
		$(a_i,S_i)\leftarrow\mathtt{Deg}\mbox{-}\mathtt{Res}\mbox{-}\mathtt{Sampling}\left(\max\left\{1,i\cdot\frac{d}{c}\right\},\frac{d}c,s\right)$
	}
	\Return{\small Uniform random neighbourhood $(a_i,S_i)$ \text{from successful runs}}
\end{algorithm}

\par In \cite{orig} it is proven that setting $s=\left\lceil\ln(n)\cdot n^{\frac1c}\right\rceil$ and using $c$ \texttt{degree-based reservoir samplers} each with a different lower bound, incremented from $\frac{d}c$ to $d$ stepping by $\frac{d}c$ each time, results in a high probability of success. The samplers are run in parallel so that the algorithm only requires a single pass of the stream. Implementing parallel running of the samplers just requires passing the same instruction to each sample before fetching the next instruction.

>>>>

\usepackage{adjustbox}
\usepackage[ruled, vlined, linesnumbered]{algorithm2e}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{changepage}
\usepackage{enumerate}
\usepackage{enumitem}
\usepackage{fancyhdr} % Header
\usepackage{float} % [H] for figure
\usepackage{graphicx}
\usepackage{subfig}
\usepackage{tikz}

@@

\renewcommand\thetable{\arabic{table}} % figure numbering strategy
\renewcommand{\thefigure}{\arabic{chapter}.\arabic{section}.\arabic{figure}}
\pagenumbering{roman}

@@

@@

% Title

\newpage
\thispagestyle{empty}

\adjustbox{padding={5pt},frame={1pt},right}{Dissertation Type: Research}

\vspace*{\fill}
\begin{center}
  \includegraphics[scale=1]{img/logo_uob_colour.PNG}\\
  \vspace*{1.0cm}
  DEPARTMENT OF COMPUTER SCIENCE\\
  \vspace*{2.0cm}
  {\LARGE Implementing and Evaluating Space Efficient Algorithms for Detecting Large Neighbourhoods in Graph Streams} \\
  \vspace*{1.0cm}
  {\Large Dominic Hutchinson}\\

  \vspace*{1.0cm}
  \rule{0.5\textwidth}{0.5pt}
  \vspace*{1.0cm}

  A dissertation submitted to the University of Bristol in accordance with the requirements of the degree of Master of Engineering in the Faculty of Engineering.

  \vspace*{1.0cm}
  \rule{0.5\textwidth}{0.5pt}
  \vspace*{1.0cm}

  May 9, 2020
\end{center}
\vspace*{\fill}

\chapter*{Declaration}
\par This dissertation is submitted to the University of Bristol in accordance with the requirements of a third year project for the degree of MEng in the Faculty of Engineering. It has not been submitted for any other degree or diploma of any examining body. Except where specifically acknowledged, it is all the work of the Author.
\par\vspace{5cm}
\par\noindent Dominic Hutchinson, Saturday $9^\text{th}$ May 2020.

\section*{Ethics}
\par This project does not require ethics approval.

@@

\par \texttt{Neighbourhood Detection} is a problem in graph theory which tasks one with finding a vertex in the graph which has a certain number of neighbours and to then return a subset of these neighbours. This is a trivial problem to solve in theory, but in practice is hard to solve within a time frame which makes the implementation practical. A major reason for this is that many approaches to solving the problem require a lot of memory and so quickly overflow a computer's RAM, slowing the execution significantly down. This common space inefficiency is becoming increasingly problematic since the move towards Big Data.

\par In this paper I discuss and implement the two algorithms presented in \cite{orig} which offer space efficient solutions for \texttt{Neighbourhood Detection} for two types of graph stream. During the implementation I discuss and evaluate several methods for improving the practical performance of these algorithms, focusing on space-efficiency in order to maximize the potential size of graphs the implementations are effective on.

\par The final implementation of the proposed algorithm for insertion-only streams proves very promising with it able to find large neighbourhoods in a couple of minutes for a graph with 30 million edges. The algorithm for insertion-deletion streams proves much more difficult to evaluate due to restrictions incurred from the implementation hash functions for $L_0$ samplers.

@@

I would like to thank Dr Christian Konrad for being an enthusiastic and engaging supervisor throughtout this project. For the help and guidance he offered throughout. In particular, I would like to thank him for being willing to supervise this project after my original supervisor was unable to do so.

@@

%TODO why big data is important
% The commercial movement towards data science and the desire to find relationships has lead to big data....
\par The movement towards Big Data has lead to ever larger data sets, which traditional offline algorithms cannot work with efficiently due to their high memory-requirements. Moreover, the size of data-sets is increasing at a much greater rate than standard RAM capacities meaning that virtual memory has to be used for these algorithms, significantly increasing their run-times.

\par The \textit{Data Stream Model} was introduced as a method for coping with these large data sets. In the data stream model instructions are received sequentially, as a \textit{`stream'}, and generally random access is not allowed. \textit{Graph Streams} are a popular implementation of this model, where instructions describe updates to the edge set of a graph and are typically ordered as a time-series. This allows for a dynamic representation of a graph, which typical \texttt{adjacency list} implementations do not. Graph streams are ideal for representing changing networks such as a social or computer network where objects can connect and then disconnect.

\par Streaming algorithms are a class of algorithms, first formalised in 1996 \cite{graphStreams}, to work within the data stream model. Streaming algorithms are typically designed to work in $polylog(n)$, where $n$ is the number of vertices in the case of graph streams. Streaming algorithms are designed to work in as few passes of the stream as possible, preferable only a single-pass is used as this means the stream can easily be extended without the algorithm having to start over. Due to the limited space streaming algorithms are allocated, they often produce an approximation of a solution in order to improve space efficiency. When implementing a streaming algorithm the trade-off between accuracy, number of passes and space-usage needs to be considered.

\par Graphs are used to represent relationships between objects and are popular in data science. A basic problem in graph theory is \texttt{High Degree Detection} where you want to find the vertices with the greatest degree. This problem can be trivially and efficiently solved by counting the number of edges incident to each vertex and then returning any vertices of sufficient degree. Solving \texttt{High Degree Detection} allows for the identification of the most influential vertices in a graph, however you cannot make any inferences about the nature of this influence. In a social network it is possible for someone to have lots of followers (\ie be of high degree), but for all these followers to be bot accounts meaning the account actually has zero influence.

@@

\par\texttt{Neighbourhood Detection} is the natural set on from \texttt{High Degree Detection}, it tasks you with finding vertices of high degree \underline{and} a subset of their neighbours. Solving \texttt{Neighbourhood Detection} allows for inferences to be made about the influence a vertex has. Extending the social network example, \texttt{Neighbourhood Detection} allows for analysis of the followers of a popular account which can be used for targeted marketing campaigns.

@@

\par The temporal-ordering of instructions in a graph stream is a hindrance for some problems, including \texttt{Neighbourhood Detection}, as it means edges likely do not arrive in the most useful order and very few assumptions can be made about what the next instruction will be. Suppose instructions were ordered by one of their incident vertex names, then we could infer when the last occurence of a vertex in the stream has occured and discard unnecessary information about that vertex.

\par In this paper I discuss two streaming algorithms which solve \texttt{Neighbourhood Detection} in $polylog(n)$ space. This is achieved by using a series of samplers in order to reduce the set of edges which need to be analysed. This means that there is a possibility of failure and this possibility increases as fewer edges are sampled, but there is a trade-off here with space efficiency.

@@

\section{Objectives}
The objectives of this project are to:
\begin{itemize}
  \item Understand the problem posed in \cite{orig} and its context in practical situtations.
	\item Implement and understand the algorithms proposed in \cite{orig}.
  \item Implement na\"ive algorithms and test their performance against the proposed.
  \item Show that na\"ive algorithms have a limit to the size of graph they can work on and that it is much lower than for the proposed algorithms.
  \item Investigate the practicalities of the samplers used in these algorithms.
  \item Discuss and test alterations to the logic and implementation of the proposed algorithms, and the subroutines they rely upon, in order to improve the performance of the proposed algorithm.
  \item Discuss the trade-offs in practicalities these alterations have, namely between speed and space usage.
	\item Use large graphs from real-world scenarios to evaluate the practicalities of these implementations on real-world data sets.
  \item Discuss how graphs from different real-world sources (\ie not social networks) might perform on these implementations and how the implementations should be changed to accommodate for them.
\end{itemize}

@@

Since the formalisation of the data streaming model in 1996 much research has been done into solving problems using this model. Here I give an overview of some of these problems and the work which has been done on them.\newline

@@

\par\noindent\textbf{Minimum Distance between two nodes}
\par An $\alpha$-\textit{spanner} subgraph $H$ of a graph $G=(V,E)$ is a subgraph where $\forall\ u,v\in V$ $d_G(u,v)\leq d_H(u,v)\leq\alpha\cdot d_G(u,v)$ where $d_G(\cdot,\cdot)$ is the length of the shortest path between the given nodes in $G$. Spanners are a space-efficient structure for evaluating the length of shortest path between any two nodes in a graph. A $(2t-1)$-spanner requires at most $\mathcal{O}(n^{1+\frac1t})$ edges need to be stored \cite{edgesOfSpanner}.
\par Baswana \cite{spanner} presents an single-pass streaming algorithm for unweighted insertion-only graph streams which runs in $\mathcal{O}(m)$ time. This is an amortised run-time of $\mathcal{O}(1)$ per edge, a natural limit for performance.\newline

\par\noindent\textbf{Random walks}
\par A random walk from $v_0\in V$ in a graph is any random sequence of vertices $v_0,v_1,v_2,\dots$ where $v_i$ is randomly selected from the neighbourhood of $v_{i-1}$. The distribution of $v_t$ can be assessed to make inferences about the structure of the graph.
\par Das Sarma et al. \cite{RandomWalks} present an algorithm for approximating the distribution of $v_t$ in the context of estmating the \textit{PageRank} of a vertex. They present an algorithm which can perform $\frac{n}{l}$ indendepent random walks of length $l$ in $o(n)$ space and $o(l)$ passes. They use this algorithm to sample random walks of different lengths in order to estimate the distribution of $v_t$ and thus estimate \textit{PageRank}.\newline

@@

\section{Structure}
Having provide some background to the problem, the structure of this project is as follows. In \texttt{Chapter 1} I discuss and define general ideas, technologies and techniques which are used throughout this project. In \texttt{Chapter 2} I discuss, implement and evaluate the algorithm proposed in \cite{orig} for insertion-only graph streams. In \texttt{Chapter 3} I discuss, implement and evaluate the algorithm proposed in \cite{orig} for insertion-deletion graph streams. And, in \texttt{Chapter 4} I review what was achieved in this project and provide some ideas for possible future work to build upon those achievements.

\section{Code}
A git repository is available containing the code for this project.\\
\texttt{https://github.com/dajhutchinson/Detecting-Large-Neighbourhoods-in-Graph-Streams}

@@

\par A graph $G=(V,E)$ is said to be a \textit{Bipartite Graph} if its vertex set $V$ can be partitioned into two disjoint subsets $A,B$ such that every edge connects a vertex in $A$ to a vertex in $B$. Formally, $\exists\ A,B\subseteq V$ st $A\cup B=V,\ A\cap B=\emptyset$ and $\forall\ (a,b)\in E$ we have that $a\in A,b\in B$. \textit{Bipartite Graphs} are denoted as $G=(A\cup B,E)$.

@@

\par The \textit{Neighbourhood} of a vertex $v$ is the set of vertices which share an edge with $N_v=\{u:(u,v)\in E\}$. The \textit{Degree} of a vertex $v$ is the size of its neighbourhood $\delta_v=|N_v|$. A vertex $v$ is \textit{$s$-sparse} if its degree is less than, or equal to, $s$. The \textit{Degree Distribution} of a graph is a distribution of the degrees of the vertices in a graph.
\par The \textit{Edge Vector} of a vertex $v$ is a vector of length $|V|$ where each index gives the weight of the edge from $v$ to the vertex associated with that index. Typically a weight of 0 is stored if no edge exists between $v$ and a specific vertex. The graphs used in this project are unweighted so the values in the edge vector are booleans describing whether, or not, an edge exists. These edge vectors can be stored as \texttt{bitstrings}.
\par \textit{Graph Streams} are an unbounded sequence of instructions $\{e_1,e_2,\dots\}$ which describe how to construct the graph. These instructions $e_i$ describe modifications to the edge set of the graph. The vertex set of the described graph is inferred from the instructions. In this project we are only concerned with two types of graph stream.
\begin{itemize}
	\item \textit{Insertion-Only Streams} where each instruction inserts a new edge to the graph. It is assumed that no edges are inserted multiple times. For the unweighted, undirected graphs used in this project insertion-only streams are a list of the edges in the graph and can be read in any order. Here the instructions have the form $e_i=(u,v)$ where $u,v\in V$ are the endpoints of the edge being inserted.
	\item \textit{Insertion-Deletion Streams} where each instruction either adds a new edge to the graph or removes an existing edge from the graph. It is assumed that no edges are inserted multiple times, nor is an edge removed if it is not currently in the graph. For the unweighted, undirected graphs used in this project instructions of insertion-deletion streams take the form $e_i=(\Delta,u,v)$ where $u,v\in V$ are the endpoints of the edge and $\Delta$ is a boolean defining whether $e_i$ is an insertion or deletion instruction. The order of instructions is clearly important for insertion-deletion streams.
\end{itemize}
Other versions of graph streams exist which allow for changes to the weights of edges and directed edges, but these are out of scope for this project.

@@

\par An \textit{$\alpha$-Approximation Algorithm} is an algorithm which returns a result within an $\alpha$ factor to the optimal solution to the problem. Algorithms which solve \texttt{Neighbourhood Detection} are $\frac1c$-approximation algorithms since they only return $d/c$ of the neighbours of a vertex which we know to have at least $d$ neighbours.

@@

\par A family of hash functions $\mathcal{H}=\{h:X\to Y\}$ is described as begin \textit{$k$-Wise Independent} if for any $k$ distinct elements of $X$ $\{v_1,\dots,v_k\}$ and any, not necessarily distinct, elements of $Y$ $\{a_1,\dots,a_k\}$ if $h$ is chosen uniformly at random from $\mathcal{H}$ then $$\mathbb{P}(h(v_1)=a_1\wedge\dots\wedge h(v_k)=a_k)=\frac1{|Y|^2}$$
\section{Technologies}
%\begin{itemize}
%	\item C++
%	\begin{itemize}
%		\item Low level
%		\item Memory manipulation
%		\item Faster
%		\item Perhaps something about why not java or python etc.
%	\end{itemize}
%	\item Standard library online
%\end{itemize}
For the implementations in this project I chose to use C++, specifically \texttt{C++11}. I initially considered using either Python or C++, as they both work in the object-orientated paradigm and I am proficient in both. I chose C++ as it allows for more control with memory management and typically has faster run-times. For this project the memory management is more important as I am looking to evaluate the space-efficiency of algorithms. The faster run-times is important when evaluating the practicalities of the algorithms.
\par I limited myself to using the \textit{C++ Standard Library} \texttt{std} as the underlying implementations of this library are well document. This was particularly important for the abstract data types I used as I could check what the underlying implementations and adjust my evaluation accordingly. When implementing randomness I used the \texttt{<random>} module of \texttt{std} and seeded the generators with the current time so that each run would have a different generator.

@@

During evaluation three performance metrics were measured: success-rate, time-taken, and space-used. These are all interconnected and come with their own trade-offs. Typically, the changes to space-used and time-taken when different strategies are implemented occur in the same direction, but at different rates. During most tests the success-rate was near perfect, but the other metrics were significantly higher as it took longer for a solution to be found.

\par In real world scenarios we are most interested analysing the neighbourhoods of the most influential vertices. For this reason $d$ is set to the maximum degree in the graph during all tests. Further, the greater the proportion of a neighbourhood returned the better the inferences made are so I focussed testing on low values of $c$, namely $c\in[2,20]$.

\par For this project I wanted some graphs generated from the real-world. I found a collection of large graph streams, from different applications, in the \texttt{Stanford Network Analysis Project} (\texttt{SNAP}) datasets collection \cite{SNAP}. I chose to use the graphs generated from social networks as I found that application most motivating, but this choice is essentially arbitrary. The graph streams from \cite{SNAP} were all insertion-only so I wrote a function in \texttt{utility.cpp} which created insertion-deletion streams from them by using a Bernoulli random variable to decide whether to delete an edge immediately after inserting it. \textbf{Table 1} provides details about the graphs used during evaluation.

@@

\par I wrote a function \texttt{utility.cpp} which took the returned vertex $v$ and set of neighbours $A$ and validated whether the final degree of $v$ was at least $d$ and that $A$ was in fact a subset of the final neighbourhood of $v$.

@@

\par The reservoir $R$ has an invariant that at any point in time it contains a uniform sample of the vertices whose degrees are known to be greater than $d_1$. This invariant is proved as \textbf{Lemma 2.1}, first I shall describe how the reservoir is maintained by controlling how vertices are inserted.

\par The first time the degree counter for a vertex $v$ surpasses $d_1$, $v$ is considered for insertion into the \textit{reservoir}. There are two cases
\begin{itemize}
	\item If the reservoir is \underline{not} full (\ie $|R|<s$) then $v$ is inserted into the reservoir.
	\item Otherwise, a Bernoulli random variable with probability $p=\frac{x}s$ is used, where $x$ is the number of vertices known to have degree at least $d_1$. If this random variable succeeds then pick, uniformly at random, a vertex $u$ currently in the reservoir and replace it with $v$. All the edges in $E$ which are incident to $u$, and no other vertex in $R$, are removed from $E$.
\end{itemize}

\noindent Whenever an edge $(u,v)$ is encountered and $u$ is in the reservoir and the number of edges stored in $E$ incident to $u$ is less than $d_2$ then $(u,v)$ is inserted into $E$.\newline

\noindent\fbox{\parbox{\textwidth}{

@@

Hence by mathematical induction, $R$ contains a uniform sample of the vertices of sufficient degree at all points in time.$\hfill\square$
}}\newline

@@

\texttt{Neighbourhood Detection} can be solved using \texttt{Degree-Based Reservoir Sampling} by setting $d_2=\frac{d}c$. The question remains as to what to set $d_1$ and $s$ to in order to achieve a high probability of success. \textbf{Algorithm 2} is the space-efficient algorithm proposed in \cite{orig} for solving \texttt{Neighbourhood Detection} for insertion-only streams.

\begin{algorithm}
	\caption{One-pass $c$-Approximation Insertion-only Streaming Algorithm for $\mathtt{Neighbourhood\ Detection}$}
	\SetKwInOut{Require}{require}
	\Require{Space $s$, degree bound $d$.}
	$s\leftarrow\lceil\log(n)\cdot n^{\frac1c}\rceil$\\
	\For{$i\in[0,c-1]$\text{\textbf{ in parallel}}} {
		$(a_i,S_i)\leftarrow$\texttt{Degree-Based Reservoir Sampling}$\left(\max\left\{1,i\cdot\frac{d}{c}\right\},\frac{d}c,s\right)$
	}
	\Return{Uniform random neighbourhood $(a_i,S_i)$ from successful runs}
\end{algorithm}

\par In \cite{orig} it is proven that setting $s=\left\lceil\ln(n)\cdot n^{\frac1c}\right\rceil$ and using $c$ \texttt{Degree-Based Reservoir Samplers} each with a different lower bound, incremented from $\frac{d}c$ to $d$ stepping by $\frac{d}c$ each time, results in a high probability of success. The samplers are run in parallel so that the algorithm only requires a single pass of the stream. Implementing parallel running of the samplers requires passing the same instruction to each sample before fetching the next instruction.

\par Each of the $c$ samplers requires $\mathcal{O}(n\log n+sd_2\log n)$ space. Since $D$ can be shared between samplers the total space is $\mathcal{O}(n\log n+c\cdot sd_2\log n)=\mathcal{O}(n\log n+n^{\frac1c}d\log^2n)$. The \textbf{Algorithm 2} does require you to know the number of vertices in the graph stream before running the algorithm, or at least the $\log$ of it. If this is unknown then it is quick to run through the whole stream and build a set of vertices in the graph. This requires $\mathcal{O}(n\log n)$ space, assuming $O(\log n)$ space is required to store a vertex. However, in many real-world scenarios this number (or a good approximation) will be known due to it being important for other tasks.
